---
title: "Workshop2_R"
format: html
---

#----workshop 2----#

## Introduction
This qmd file covers the material from the second workshop for the R in marine science module. 
here we are covering:
- Tidying data
- pivoting wider and longer
- spearating and uniting variables
- dealing with strings and dates
- joining separate date frames/tibbles
- Handling missing data


## Tidying data
How we make our dataset tidy is by following three interrelated rules. 
Each variable must have its own column.
Each observation must have its own row.
Each value must have its own cell.

```{r}
#loading library
library(tidyverse)
library(palmerpenguins)
library(dplyr)


```
Understanding whether your data frame structure is optimal (or tidy) is a fundamental skill for a data scientist

Step 1 = tidy and understand what the data is and means
step 2 =  resolve one of the two common problems with untidy data. These are:
    One variable is spread across multiple columns
    One observation is scattered across multiple rows
    
## Lengthening datasets
-using pivot_longer() to lengthen dataset
example of wide format:
billboard
#> # A tibble: 317 × 79
#>   artist       track               date.entered   wk1   wk2   wk3   wk4   wk5
#>   <chr>        <chr>               <date>       <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 2 Pac        Baby Don't Cry (Ke... 2000-02-26      87    82    72    77    87
#> 2 2Ge+her      The Hardest Part O... 2000-09-02      91    87    92    NA    NA
#> 3 3 Doors Down Kryptonite          2000-04-08      81    70    68    67    66
#> 4 3 Doors Down Loser               2000-10-21      76    76    72    69    67
#> 5 504 Boyz     Wobble Wobble       2000-04-15      57    34    25    17    17
#> 6 98^0         Give Me Just One N... 2000-08-19      51    39    34    26    26
#> # ℹ 311 more rows
#> # ℹ 71 more variables: wk6 <dbl>, wk7 <dbl>, wk8 <dbl>, wk9 <dbl>, ...



```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )
```

Note that in the code "week" and "rank" are quoted because they are new variables that we are creating, they don’t exist yet in the data when we run the pivot_longer() call.

-cols which specifies the columns you want to pivot (the ones that aren’t variables). Note: you could either use !c(artist, track, date.entered) OR starts_with('wk') because the cols argument uses the same syntax as select().

-names_to which names the variable stored in the column names. We chose to name that variable week.

-values_to which names the variable stored in the cell values that we named rank

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
#> # A tibble: 5,307 × 5
#>   artist track                   date.entered week   rank
#>   <chr>  <chr>                   <date>       <chr> <dbl>
#> 1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#> 2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#> 3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#> 4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#> 5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#> 6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#> # ℹ 5,301 more rows
```
adding the values_drop_na = TRUE argument to the pivot_longer() call drops all rows where the rank variable is NA. This is useful because we only want to keep rows where there is a rank value, and not rows where there is no rank value (NA).

mutate() and parse_number() are used to convert the week variable from a character variable (e.g. "wk1") to a numeric variable (e.g. 1). This is done by first using parse_number() to extract the numeric part of the week variable, and then using mutate() to create a new variable called week_num that contains the numeric values.

```{r}
billboard_longer <- billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  ) |> 
  mutate(
    week = parse_number(week)
  )

billboard_longer

```

#Example exercise:
```{r}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
```

we want our new (tidy) dataset to have three variables: 
id (which already exists)
measurement (the column names) 
value (the cell values)

```{r}

df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```

## Widening datasets
-using pivot_wider() to widen dataset -  allows us to handle an observation if it is scattered across multiple rows.

```{r}
cms_patient_experience
```

We can see the complete set of values for measure_cd and measure_title by using distinct():

```{r}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)
```

pivot_wider() has the opposite interface to pivot_longer(): instead of choosing new column names, we need to provide the existing columns that define the values (values_from) and the column name (names_from):

```{r}
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
```

need to tell pivot_wider() which column or columns have values that uniquely identify each row; in this case those are the variables starting with "org":

```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )

```

Pivoting wider - example

```{r}
df1  <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```

take the names from the measurement column using the names_from() argument and the values from the value column using the values_from() argument:

```{r}
df1 |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```
```{r}
df1 |> 
  distinct(measurement) |> 
  pull()
```
```{r}
df1 |> 
  select(-measurement, -value) |> 
  distinct()

```

pivot_wider() then combines these results to generate an empty dataframe:

```{r}
df1 |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)

```


## Pivoting Exercises using Palmer Penguins

```{r}
library(palmerpenguins)

# Create a long version of the penguins dataset

penguins_long <- penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# View the result
head(penguins_long)

```
Visulising the data:
```{r}
penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )

```

Widening data with pivot_wider()

first calculate the mean body mass of each speceis for each island using group_by() and summarise():

```{r}
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary)

```

use pivot_wider() to pull the island names up into columns, creating a clean matrix of species by island.
```{r}
mass_summary_wide <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )


head(mass_summary_wide)

```
##  Separating and uniting data tables

```{r}
table3

```
 need to split the rate column up into two variables: 1) cases and 2) population. separate() will take the name of the column we want to split and the names of the columns we want it split into
 
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"))

```
```{r}
#rewrite the code above as to change the default although this is what the default is:
table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")


```
```{r}
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE) # this converts the default character variables into numeric variables

```
 
```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2) # seperates the year into 2 variables - centrury and year

```
 
Using unite():
unite() takes a data frame, the name of the new variable and a set of columns to combine using dplyr::select().

we are going to rejoin the centruy and year variables back together using unite() and we will call the new variable year again:

```{r}
table5 %>% 
  unite(new, century, year, sep = "") # need to add sep ="" because we don’t want any separator (the default is to add an underscore _)
```


## Wrangling strings and dates

Standardising text with stringr

```{r}
library(tidyverse)

# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
     # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean),
    # 3. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_"),
   
  )

print(clean_sites)

```
Mastering time with lubridate

```{r}
library(lubridate)

# Parsing different date formats
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")

# R now recognizes these as identical Date objects
date_1 == date_2

```
If your dataset contains a date formatted as "25/12/2026", the day comes first, then the month, then the year. Therefore, you use the dmy() function. If the date is "2026-12-25", you use ymd().


For high-frequency marine data, such as acoustic telemetry or CTD profiles, we often need to parse exact timestamps. We do this by appending _hms (hours, minutes, seconds) to our functions.:

```{r}
# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)


```

```{r}
ggplot(sensor_clean, aes(x = true_time, y = temperature)) +
  geom_line() +
  labs(title = "Temperature over Time", x = "Time", y = "Temperature (°C)") +
  theme_minimal()
```


## Relational data: Joining tables

```{r}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB"),
  species = c("Trout", "Snapper", "Trout", "Cod"),
  count = c(5, 2, 1, 3)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12)
)

```

