---
title: "module 2 - R in marine science"
author: "Ben Barber"
date: "2026-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Housekeeping
## Inventory every active object currently residing in session RAM
objects()
## Purge global environment
rm(list = ls())
# Confirm that global session memory now completely vacant
objects()


## Ingesting data into R ##

practice import of datasets:
```{r}
# Load the primary data science framework and Excel import library
library(tidyverse)
library(readxl)

# Practice Import A: Loading a standard comma-separated plain text file
benthic_cover <- read_csv(here::here("data/reef_cover_log.csv"))

# Practice Import B: Parsing a tab-separated telemetry instrument array string
acoustic_stream <- read_tsv(here::here("data/acoustic_telemetry_stream.txt"))

# Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet
fisheries_annual <- read_excel(here::here("data/fish_catch_data.xlsx"), sheet = "Commercial_2026")

# Read in mangrove_data
mangrove_data <- read_csv(file = here::here("data/mangrove_survey_raw.csv"))

# Use args within read_csv to skip headers and declare missing flags
mangrove_data <- read_csv(
  here::here("data/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA

```


## Data frame architectures: Tibbles versus legacy tables ##

```{r}
# Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)

```

compare this:
```{r}
# Print the old-style dataframe structure to view
print(benthic_cover_df)
# And compare with tibble alternative
print(benthic_cover)
 
```
Observe your console feedback. The tibble structure immediately tells you its dimensions and highlights variable configurations right below the column headers. This constant visual validation is your primary defense against typing errors.


## Wrangling out ecological signals using Palmer Penguins ##

```{r}
# Install the data package (execute this command once in your console pane and then delete!)
# install.packages("palmerpenguins")

# Load the package data into active memory
library(palmerpenguins)
library(tidyverse)
data("penguins")

# Examine the structure of the dataset - always do this when loading a new dataset!
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version

```
```{r}
# Generate an exploratory summary matrix
summary(penguins)

```

The summary() printout is an immediate diagnostic tool. It reports the presence of explicit missing indicators (NA) within individual biological metrics. For instance, if sex or body_mass_g lists several NA counts, it flags that certain individual penguins escaped the researchers during that day's field sampling!


## Foundational grammar: Slicing, filtering, sorting, and transforming ##
Isolating attributes with select()

```{r}
# Vertically slice specific morphometric variables by explicit name
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)

# Discard logistics tracking attributes while preserving everything else using the minus sign
clean_scientific_fields <- select(penguins, -year)

```
## Sifting rows with filter()

```{r}
# Isolate observations belonging to a single categorical target group
adelie_cohort <- filter(penguins, species == "Adelie")

# Sift out individuals using continuous numerical boundary thresholds
# Preserves only large penguins whose mass exceeds 4500 grams
heavy_penguins <- filter(penguins, body_mass_g > 4500)

# Combine multiple conditional parameters across separate attributes
# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

# Sift records matching multiple targeting flags within an explicit set
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))

```

## Ordering sequences with arrange()

The arrange() verb alters the sorting configuration of rows within your data frame without changing individual cell values. This is essential for inspecting size hierarchies or chronologically ordering long-term environmental observation lines:

```{r}
# Sort penguins by ascending body mass (Default setting: Smallest mass first)
lightest_first <- arrange(penguins, body_mass_g)

# Sort penguins in descending sequence using the desc() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))

# Execute nested sorting criteria: Group by species, then sort by descending bill length
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))

```

##  Introducing the Pipe (|>)
The syntax for the native R pipe is |>. (Note there is also a pipe operator built into the tidyverse package magrittr, which is very common (%>% )


Example: Instead of:
penguins_subset <- mutate(penguins, bill_ratio = bill_length_mm / bill_depth_mm)
penguins_final <- filter(penguins_subset, species == "Adelie")

You can simply write:
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")



## Computing new attributes with mutate()

```{r}
# Calculate a new morphological ratio in our environment
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )

# View your newly engineered variables appended to the far-right columns
glimpse(penguin_ratios)

```

## Data aggregation and ecological summarisation

```{r}
# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)

```
This is an essential behavioral design feature of R known as the Missing Value Trap. If a column contains even a single missing observation flag (NA), any mathematical aggregation function (such as mean(), sd(), or sum()) will automatically return NA to protect you from miscalculating metrics on incomplete data. To override this safely, you must explicitly declare that missing cells should be dropped during calculation. Here we’ll now use the magrittr pipe to wrangle our data, avoiding the missing value trap.


```{r}
# Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)

```

##  Integrating data grammar with visual diagnostics in qmd
Wrangling and Plotting in Parallel
Challenge: Visualising Body Size and Shape 
Using your loaded data, create a code chunk in your .qmd that performs the following two tasks simultaneously:
Summary Table: Pipe your data to group_by() and summarise() to calculate the mean body mass for each species and island.
Visualisation: Pipe the same dataset into ggplot() to create a boxplot of body_mass_g by species, with the island variable mapped to fill or facet_wrap().

```{r}
penguins |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = species, y = mean_mass, fill = island)) +
  geom_col()+
  theme_minimal()
```


Piping Directly to Visualisation
Example: Visualising Mean Trends with Uncertainty Instead of creating a summary table object first, you can pipe your grouped data straight into a plot and use geom_errorbar() to represent the spread of your data:

```{r}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```


Challenge: 
In your .qmd template, try creating a new plot. Experiment with swapping sd_mass (standard deviation) for standard error—you will need to divide your sd by the square root of n (sd / sqrt(n)). See how that changes your visual interpretation of the data uncertainty. Present the summary table alongside the associated figure and write down your thoughts on the underlying ecological story that the data are telling.

```{r}
# Calculate standard error and plot with error bars
# Pipe directly from aggregation to plotting with error bars
SE_mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - (sd_mass / sqrt(n)), 
                    ymax = mean_mass + (sd_mass / sqrt(n))), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard error",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

print(SE_mass_compare_plot)
```


## Ecological Story

The two plots present different aspects of the uncertainty surrounding penguin body mass. The first plot, using **standard deviation (SD)**, illustrates the *spread of individual body masses* within each species. It shows the inherent variability among individual penguins of the same species. For example, Adélie penguins show a notable range of body masses, indicating that individual penguins can vary quite a bit from the species mean. This measure is crucial for understanding the biological diversity and range of traits within a population.

In contrast, the second plot, using **standard error (SE)**, reflects the *precision of our estimate of the mean body mass* for each species. Because the standard error accounts for the sample size (`n`), the error bars are considerably narrower than those representing standard deviation. This smaller uncertainty suggests that we are quite confident about the calculated average body mass for each species, given the number of penguins sampled. The overlap in SE bars between Chinstrap and Gentoo is less pronounced than with SD, but still present, especially between Adélie and Chinstrap. This suggests that while there is individual variation (SD), our estimate of the average mass (SE) is more precise, potentially indicating statistically significant differences in mean mass between some species, particularly between Gentoo and the other two.

The key takeaway for the ecological story is that while penguins exhibit natural individual variability (SD), the large sample sizes collected for each species allow for a precise estimation of their average body mass (SE). This precision allows us to discern clearer differences in average size between species, such as the significantly larger body mass of Gentoo penguins compared to Adélie and Chinstrap, while acknowledging the broader range of sizes individuals can reach.

#saving data
```{r}

library(readr)
# 1. Exporting our collapsed summary table as a universal flat text file
write_csv(biological_signal, "../Outputs/tables/penguin_species_mass_summary.csv")
getwd()

# 2. Saving our cleaned morphological cohort table as a native R binary file
saveRDS(morphology_metrics, "../outputs/tables/morphology_metrics.rds")

```
```{r}
ggsave("../Outputs/figures/SE_mass_compare_plot.png", 
       plot = mass_compare_plot, 
       width = 120, height = 120, 
       units = "mm", dpi = 300)

```

