---
title: "module 2 - R in marine science"
author: "Ben Barber"
date: "2026-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Housekeeping
## Inventory every active object currently residing in session RAM
objects()
## Purge global environment
rm(list = ls())
# Confirm that global session memory now completely vacant
objects()


## Ingesting data into R ##

practice import of datasets:
```{r}
# Load the primary data science framework and Excel import library
library(tidyverse)
library(readxl)

# Practice Import A: Loading a standard comma-separated plain text file
benthic_cover <- read_csv(here::here("data/reef_cover_log.csv"))

# Practice Import B: Parsing a tab-separated telemetry instrument array string
acoustic_stream <- read_tsv(here::here("data/acoustic_telemetry_stream.txt"))

# Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet
fisheries_annual <- read_excel(here::here("data/fish_catch_data.xlsx"), sheet = "Commercial_2026")

# Read in mangrove_data
mangrove_data <- read_csv(file = here::here("data/mangrove_survey_raw.csv"))

# Use args within read_csv to skip headers and declare missing flags
mangrove_data <- read_csv(
  here::here("data/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA

```


## Data frame architectures: Tibbles versus legacy tables ##

```{r}
# Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)

```

compare this:
```{r}
# Print the old-style dataframe structure to view
print(benthic_cover_df)
# And compare with tibble alternative
print(benthic_cover)
 
```
Observe your console feedback. The tibble structure immediately tells you its dimensions and highlights variable configurations right below the column headers. This constant visual validation is your primary defense against typing errors.


## Wrangling out ecological signals using Palmer Penguins ##

```{r}
# Install the data package (execute this command once in your console pane and then delete!)
# install.packages("palmerpenguins")

# Load the package data into active memory
library(palmerpenguins)
library(tidyverse)
data("penguins")

# Examine the structure of the dataset - always do this when loading a new dataset!
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version

```
```{r}
# Generate an exploratory summary matrix
summary(penguins)

```

The summary() printout is an immediate diagnostic tool. It reports the presence of explicit missing indicators (NA) within individual biological metrics. For instance, if sex or body_mass_g lists several NA counts, it flags that certain individual penguins escaped the researchers during that day's field sampling!


## Foundational grammar: Slicing, filtering, sorting, and transforming ##
Isolating attributes with select()

```{r}
# Vertically slice specific morphometric variables by explicit name
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)

# Discard logistics tracking attributes while preserving everything else using the minus sign
clean_scientific_fields <- select(penguins, -year)

```
## Sifting rows with filter()

```{r}
# Isolate observations belonging to a single categorical target group
adelie_cohort <- filter(penguins, species == "Adelie")

# Sift out individuals using continuous numerical boundary thresholds
# Preserves only large penguins whose mass exceeds 4500 grams
heavy_penguins <- filter(penguins, body_mass_g > 4500)

# Combine multiple conditional parameters across separate attributes
# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

# Sift records matching multiple targeting flags within an explicit set
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))

```

## Ordering sequences with arrange()

The arrange() verb alters the sorting configuration of rows within your data frame without changing individual cell values. This is essential for inspecting size hierarchies or chronologically ordering long-term environmental observation lines:

```{r}
# Sort penguins by ascending body mass (Default setting: Smallest mass first)
lightest_first <- arrange(penguins, body_mass_g)

# Sort penguins in descending sequence using the desc() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))

# Execute nested sorting criteria: Group by species, then sort by descending bill length
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))

```

